---
title: "pca_part_2"
author: "Biswajit Sahoo"
date: "December 21, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
This post is Part-II of a three part series post on PCA. Other parts of the series can be found at the links below. 

* [Part-I: Basic Theory of PCA](https://biswajitsahoo1111.wordpress.com/2018/12/29/principal-component-analysis-part-i/)
* [Part-III: Reproducing results of a published paper on PCA](https://biswajitsahoo1111.wordpress.com/2018/12/29/principal-component-analysis-part-iii/)

In this post we will first apply built in commands to obtain results and then we will show how the same results can be obtained without using built-in commands. By this post, our aim is not to advocate the use of non-built-in functions. Rather, in our opinion, it enhances understanding by knowing what happens under the hood when a built-in function is called. In actual applications, readers should always use built functions as they are robust(almost always) and tested for efficiency. 

In this post readers can find code snippets for R. Equivalent [MATLAB codes](https://github.com/biswajitsahoo1111/PCA/blob/master/pca_part_II_MATLAB_codes.pdf) for the same can be obtained from this [link](https://github.com/biswajitsahoo1111/PCA/blob/master/pca_part_II_MATLAB_codes.pdf). 

We will use French food data form reference [2]. Refer to the paper to know about the original source of the data. We will apply different methods to this data and compare the result.

### Load Data

```{r}
#Load abdi food data
(food = read.csv("pca_abdi_food.csv",header= T))
# Centerd data matrix
cent_food = scale(food[,3:9],scale = F)
# Scaled data matrix
scale_food = scale(food[,3:9],scale = T)
```

## Covariance PCA

### Using built-in function

```{r}
# Using built-in function
pca_food_cov = prcomp(food[,3:9],scale = F)
# Loading scores (we have printed only four columns out of seven)
(round(pca_food_cov$rotation[,1:4],2))
# Factor score (we have printed only four PCs out of seven)
```
We have printed only four columns of loading scores out of seven.
```{r}
(round(pca_food_cov$x[,1:4],2))
```
We have printed only four principal components out of seven.
```{r}
# Variances using built-in function
(round(pca_food_cov$sdev^2,2))
# Total variance
(sum(round(pca_food_cov$sdev^2,2)))
```

## Comparison of variance before and after transformation

```{r}
# Total variance before transformation
sum(diag(cov(food[,3:9])))
# Total variance after transformation
sum(diag(cov(pca_food_cov$x)))
```
Another important observation is to see how variance of each variable before transformation changes into variance of principal components. Note that total variance in this process remains same as seen from above codes.
```{r}
# Variance along variables before transformation
round(diag(cov(food[,3:9])),2)
```
Note that calculation of variance is unaffected by centering data matrix. So variance of original data matrix as well as centered data matrix is same. Check it for yourself. Now see how PCA transforms these variance.
```{r}
# Variance along principal compoennts
round(diag(cov(pca_food_cov$x)),2)
# We can obtain the same result using built-in fucntion
round(pca_food_cov$sdev^2,2)
```

### Performing covariance PCA manually using SVD

```{r}
svd_food_cov = svd(cent_food)
# Loading scores
round(svd_food_cov$v[,1:4],2) # We have printed only four columns
# Factor scores
round((cent_food %*% svd_food_cov$v)[,1:4],2) # only 4 columns printed
# Variance of principal components
round(svd_food_cov$d^2/11,2)
```
Our data matrix contains 12 data points. So to find variance of principal components we have to divide the square of the diagonal matrix by 11. To know the theory behind it, refer [Part-I](https://biswajitsahoo1111.wordpress.com/2018/12/29/principal-component-analysis-part-i/)

### Performing covariance PCA using Eigen-decomoposition(Not recommended)

This procedure is not recommended because forming a covariance matrix is computationally not efficient for large matrices if data matrix contains smaller entries. So doing eigen analysis on covariance matrix may give erroneous results. However, for our example we can use it to obtain results.
```{r}
eigen_food_cov = eigen(cov(cent_food))
# Loading scores
round(eigen_food_cov$vectors[,1:4],2)
# Factor scores
round((cent_food %*% eigen_food_cov$vectors)[,1:4],2)
# Variance along principal components
round(eigen_food_cov$values,2)
```
Instead of using the 'cov()' command to find the covariance matrix manually and perform its eigen analysis.
```{r}
cov_matrix_manual_food = (1/11)*t(cent_food) %*% cent_food
eigen_food_new = eigen(cov_matrix_manual_food)
# Loading scores
round(eigen_food_new$vectors[,1:4],2)
# Variance along principal components
round(eigen_food_new$values,2)
```
There are also different ways to find total variance of the data matrix. We will explore some of the options.
```{r}
# Total varaiance before transformation
sum(diag(cov(cent_food)))
```
Note that total variance is invariant to translations. So calculating the total variance on raw data will also give the same answer. Check it to convince yourself.

## Correlation PCA

When PCA is performed on a scaled data matrix (each variable is centered as well as variance of each variable is one), it is called correlation PCA. Before discussing correlation PCA we will take some time to see different ways in which we can obtain correlation matrix. 

### Different ways to obtain correlation matrix.

```{r}
# Using built-in command
round(cor(food[,3:9]),2)[,1:4] # We have printed only four columns
# manually
round((1/11)*t(scale_food) %*% scale_food,2)[,1:4]
```

## Performing correlation PCA using built-in function
```{r}
pca_food_cor = prcomp(food[,3:9],scale = T)
# Loading scores
round(pca_food_cor$rotation[,1:4],2) # Printed only four
# Factor scores
round(pca_food_cor$x[,1:4],2)
# Variances along principal componentes
round(pca_food_cor$sdev^2,2)
# Sum of vairances
sum(pca_food_cor$sdev^2)
```

## Comparison of variance before and after transformation

```{r}
# Total variance before transformation
sum(diag(cov(scale_food)))
# Total variance after transformation
sum(diag(cov(pca_food_cor$x)))
```
Another important observation is to see how variance of each variable before transformation changes into variance of principal components. Note that total variance in this process remains same as seen from above codes.
```{r}
# Variance along variables before transformation
round(diag(cov(scale_food)),2)
```
This is obvious as we have scaled the matrix. Now see how PCA transforms these variance.
```{r}
# Variance along principal compoennts
round(diag(cov(pca_food_cor$x)),2)
# We can obtain the same result using built-in fucntion
round(pca_food_cor$sdev^2,2)
```

### Performing correlation PCA manually using SVD

```{r}
svd_food_cor = svd(scale_food)
# Loading scores
round(svd_food_cor$v[,1:4],2)
# Factor scores
round((scale_food %*% svd_food_cor$v)[,1:4],2)
# Variance along each principcal component
round(svd_food_cor$d^2/11,2)
# Sum of variances
sum(svd_food_cor$d^2/11)
```
Again we have to divide by 11 to get eigenvalues of correlation matrix. Check the formulation of correlation matrix using scaled data matrix to convince yourself. 

### Using eigen-decomposition (Not Recommended)

```{r}
eigen_food_cor = eigen(cor(food[,3:9]))
# Loading scores
round(eigen_food_cor$vectors)
# Factor scores
round((scale_food %*% eigen_food_cor$vectors)[,1:4],2)
# Variances along each principal component
round(eigen_food_cor$values,2)
```

I hope this post would help clear some of the confusions that a beginner might have while encountering PCA for the first time. Please comment below if you find any errors.

## References

1. I.T. Jolliffe, Principal component analysis, 2nd ed, Springer, New York,2002.
2. Abdi, H., & Williams, L. J. (2010). Principal component analysis. Wiley interdisciplinary reviews: computational statistics, 2(4), 433-459. 
